---
title: "Relationships between two variables"
subtitle: "REFERENCE SHEET"
format: 
  pdf:
    toc: true
    toc-depth: 5
    include-in-header: 
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
          commandchars=\\\{\},
          breaklines, breaknonspaceingroup, breakanywhere
        }
execute:
  message: false
  warning: false
  echo: false
  fig-width: 8
  fig-height: 6
---

```{r}
#| label: packages

library(tidyverse)
library(janitor)
library(ggrepel)
library(cowplot)
library(here)
```

```{r}
#| label: data-and-cleaning

birds <- read_csv("https://raw.githubusercontent.com/ENVS-193DD/datasets/refs/heads/main/birds/birds.csv")

weather <- read_csv("https://raw.githubusercontent.com/ENVS-193DD/datasets/refs/heads/main/weather/4310193.csv")

weather_clean <- weather |> 
  # clean column names
  clean_names() |> 
  # extract month
  mutate(month = month(date)) |> 
  # calculate total montly precipitation and mean monthly max temp
  group_by(month) |> 
  mutate(total_monthly_precip_in = sum(prcp, na.rm = TRUE),
         mean_monthly_tmax_f = mean(tmax, na.rm = TRUE)) |> 
  ungroup() |> 
  # select columns of interest
  select(date, total_monthly_precip_in, mean_monthly_tmax_f) |> 
  # filter to only include unique values
  distinct()
```

```{r}
#| label: wrangling-all-shorebirds

# creating new object from birds
shorebirds <- birds |> 
  # filter to only include shorebords and observations in 2025 water year
  filter(e_bird_group == "Shorebirds") |> 
  # calculate total counts per species for a survey
  group_by(observation_date, species) |> 
  summarize(total = sum(count)) |> 
  ungroup() |> 
  # taking out species only seen once
  filter(!(species %in% c("Peep sp.", "Sandpiper sp.", "Dowitcher sp.", 
                          "Wilson's Snipe", "Red-necked Phalarope"))) |> 
  # fill in any missing observation date and species combinations with 0
  complete(observation_date, species,
           fill = list(total = 0)) |> 
  # add column to denote family
  mutate(family = case_when(
    species %in% c("Black-necked Stilt", "American Avocet") ~ "Recurvirostridae",
    species %in% c("Greater Yellowlegs", "Lesser Yellowlegs", 
                   "Killdeer", "Long-billed Curlew",
                   "Least Sandpiper", "Long-billed Dowitcher", 
                   "Western Sandpiper", "Pectoral Sandpiper", 
                   "Dunlin", "Sanderling", "Solitary Sandpiper",
                   "Spotted Sandpiper", "Whimbrel",
                   "Willet") ~ "Scolopacidae",
    species %in% c("Semipalmated Plover", "Western Snowy Plover") ~ "Charadriidae"
  )) |> 
  # add a column to direct label lines
  group_by(species) |> 
  mutate(species_label = case_when(
    observation_date == min(observation_date) ~ species
  )) |> 
  ungroup() |> 
  mutate(species = fct_reorder(species, total, .fun = median, .desc = TRUE)) |> 
  # join with weather data
  left_join(weather_clean, by = c("observation_date" = "date"))
```

```{r}
#| label: wrangling-2025-shorebirds

shorebirds_wy2025 <- shorebirds |> 
  filter(between(observation_date, as_date("2024-10-01"), as_date("2025-09-30"))) |> 
  # add a column to direct label lines
  group_by(species) |> 
  mutate(species_label = case_when(
    observation_date == min(observation_date) ~ species
  )) |> 
  # filter out any species that don't appear in the 2025 water year
  filter(sum(total) != 0) |> 
  # ungroup data frame
  ungroup() 
```

```{r}
#| label: plot-aesthetics

colors <- c(
  "#0077BB",
  "#33BBEE",
  "#009988",
  "#EE7733",
  "#CC3311",
  "#EE3377",
  "#AA4499"
)

theme_set(
  theme_bw() +
    theme(text = element_text(size = 14))
)
```

\newpage

# 1. Timeseries plots

## a. Basic plot

Issues:

-   legend takes up space
-   hard to distinguish trends between species

```{r}
#| label: basic-timeseries

# base layer
basic_plot <- ggplot(data = shorebirds_wy2025,
                                # x-axis: survey date
                                # y-axis: total counts
                                mapping = aes(x = observation_date,
                                              y = total,
                                              color = species)) +
  # basic geometries -----
  # points to represent counts at each survey
  geom_point() +
  # lines to connect points
  geom_line()

basic_plot
```

\newpage

## b. Direct labelling

Improvements:

-   directly labelled species (getting rid of legend)
-   highlighting major events in time

```{r}
#| fig-height: 8
#| fig-width: 10

# base layer
shorebirds_directlabel <- ggplot(data = shorebirds_wy2025,
                                # x-axis: survey date
                                # y-axis: total counts
                                mapping = aes(x = observation_date,
                                              y = total,
                                              color = species)) +
  # basic geometries -----
  # points to represent counts at each survey
  geom_point() +
  # lines to connect points
  geom_line() +
  
  # axis appearance -----
  # adjusting appearance of x-axis
  scale_x_date(
    # expanding axis
    expand = expansion(mult = c(0.3, 0.1)),
    # setting limits (no upper bound)
    limits = c(as_date("2024-10-01"), NA),
    # setting breaks so that date labels appear every month
    breaks = as_date(c("2024-10-01", "2024-11-01", "2024-12-01",
                       "2025-01-01", "2025-02-01", "2025-03-01",
                       "2025-04-01", "2025-05-01", "2025-06-01",
                       "2025-07-01", "2025-08-01", "2025-09-01")),
    # label dates with month abbreviations
    date_labels = "%b") +
  # set y-axis expansion
  scale_y_continuous(expand = expansion(mult = c(0.01, 0.1))) +
  
  # species text labels -----
  geom_text_repel(
    # new aesthetics for label text
    mapping = aes(label = species_label),
    # setting random seed 
    # (play around with this number until you get an arrangement you like)
    seed = 100,
    # right align labels
    hjust = 1,
    # repel in y direction only
    direction = "y",
    # linetype
    segment.linetype = 3,
    # text
    size = 5,
    # nudge starting point of label to the left along x-axis and up along y-axis
    nudge_x = -1,
    nudge_y = 1,
    # lengthen line length
    segment.size = 0.5,
    # amount of repulsion between text labels
    force = 0.5,
    # set the x-axis position for where the labels should start
    xlim  = as_date("2024-10-01"),
    # remove text from the plot legend
    show.legend = FALSE) +
  
  # vertical lines for annotations -----
  # add vertical line for major rain event in January
  geom_vline(xintercept = as_date("2025-01-26"),
             linetype = "dashed") +
  # text label (from cowplot package)
  draw_label("First major rain event\n(3.40 in)",
             # x- and y-axis position
             x = as_date("2025-01-20"),   
             y = 85,
             # right alignment
             hjust = 1,
             # text size
             size = 14) +
  # add vertical line for major rain event in March
  geom_vline(xintercept = as_date("2025-03-12"),
             linetype = "dashed") +
  # text label
  draw_label("Last major rain event\n(1.11 in)",
             # x- and y-axis position
             x = as_date("2025-03-17"),   
             y = 85,
             # left alignment
             hjust = 0,
             # text size
             size = 14) +
  
  # other plot appearance customization -----
  # set custom colors
  scale_color_manual(values = colors) +
  # relabelling x- and y-axes
  labs(x = "Date",
       y = "Total counts") +
  # theme adjustments
  theme(legend.position = "none",
        panel.grid = element_blank(),
        legend.background = element_blank())


shorebirds_directlabel
```

\newpage

## c. Filled areas

Improvements:

-   easier to compare areas than points/lines
-   legend inside plot area (takes up less space)

```{r}
shorebirds_area <- ggplot(data = shorebirds_wy2025,
                          mapping = aes(x = observation_date,
                                        y = total,
                                        fill = species)) +
  # first layer: filled areas representing species counts
  geom_area() +
  
  # axis appearance -----
  # adjusting appearance of x-axis
  scale_x_date(
     # setting breaks so that date labels appear every month
     breaks = as_date(c("2024-10-01", "2024-11-01", "2024-12-01",
                        "2025-01-01", "2025-02-01", "2025-03-01",
                        "2025-04-01", "2025-05-01", "2025-06-01",
                        "2025-07-01", "2025-08-01", "2025-09-01")),
     # label dates with month abbreviations
     date_labels = "%b",
     expand = c(0, 0)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.01))) +
    
  # other plot appearance customization -----
  # set custom colors
  scale_fill_manual(values = colors) +
  # relabelling x- and y-axes
  labs(x = "Date",
       y = "Total counts",
       fill = "Species") +
  # theme adjustments
  theme(panel.grid = element_blank(),
        legend.position = c(0.8, 0.8)) 


shorebirds_area
```

\newpage

## d. Highlighting lines in multiple plots

Improvements:

-   showing each species individually while keeping other species to compare
-   easier to see trends for each species

```{r}
#| label: highlights
#| fig-width: 11

# creating new data frame for highlighted plot
highlight_df <- shorebirds_wy2025 |> 
  # creating new column that is a duplicate of existing species column
  mutate(species2 = species)

shorebirds_highlight <- ggplot(mapping = aes(x = observation_date, 
                                             y = total)) +
  # first layer: grey lines for all species counts
  geom_line(data = highlight_df |> select(-species), 
            mapping = aes(group = species2), 
            color = "grey", 
            linewidth = 0.5, 
            alpha = 0.5) +
  # second layer: colored lines for individual species counts
  geom_line(data = highlight_df, 
            mapping = aes(color = species),
            linewidth = 1) +
  # different panels for each species
  facet_wrap(~species) +
  
  # axis appearance -----
  # adjusting appearance of x-axis
  scale_x_date(
    # setting breaks so that date labels appear every month
    breaks = as_date(c("2024-10-01", "2024-11-01", "2024-12-01",
                       "2025-01-01", "2025-02-01", "2025-03-01",
                       "2025-04-01", "2025-05-01", "2025-06-01",
                       "2025-07-01", "2025-08-01", "2025-09-01")),
    # label dates with month abbreviations
    date_labels = "%b") +
  # set y-axis expansion
  scale_y_continuous(expand = expansion(mult = c(0.01, 0.1))) +
  
  # other plot appearance customization -----
  # set custom colors
  scale_color_manual(values = colors) +
    # relabelling x- and y-axes
    labs(x = "Date",
         y = "Total counts") +
    # theme adjustments
    theme(legend.position = "none",
          panel.grid = element_blank(),
          legend.background = element_blank(),
          strip.background = element_rect(color = "#000000",
                                          fill = "#FFFFFF"))

shorebirds_highlight
```

\newpage

# 2. Relationships between variables

## a. Total shorebird counts as a function of temperature

Each point represents a survey. The y-axis represents the total number of shorebirds counted in a survey, and the x-axis represents the mean monthly maximum temperature for the month of that survey.

```{r}
#| label: totals-temp-fig
#| fig-width: 8
#| fig-height: 6

# creating new object
shorebirds_totals <- shorebirds |> 
  # calculating total counts for all shorebirds
  group_by(observation_date) |> 
  summarize(total_allspp = sum(total)) |> 
  ungroup() |> 
  # joining with weather data
  left_join(weather_clean, by = c("observation_date" = "date"))

# plotting temp and total counts
temp_count_scatterplot <- ggplot(data = shorebirds_totals,
       # x-axis: mean maximum temperature
       # y-axis: total counts
       mapping = aes(x = mean_monthly_tmax_f,
                     y = total_allspp)) +
  # points to represent mean monthly temp and total counts of shorebirds
  # for each survey
  geom_point(shape = 21) +
  # relabelling axes, adding title
  labs(x = "Mean monthly maximum temperature (\U00B0 F)",
       y = "Total counts",
       title = "All shorebird species") +
  # moving title
  theme(plot.title.position = "plot")

# displaying plot
temp_count_scatterplot
```

\newpage

## b. Total species counts as a function of temperature

Each point represents a survey. The y-axis represents the total number counted in a survey for a given species, and the x-axis represents the mean monthly maximum temperature for the month of that survey.

```{r}
#| label: species-temp-fig
#| fig-height: 8

temp_count_spp <- shorebirds |> 
  # filtering for top 5 species
  filter(species %in% c("Killdeer", "Least Sandpiper", "Black-necked Stilt", 
                        "Greater Yellowlegs", "Western Sandpiper")) |> 
  # x-axis: temp, y: total counts for each species
  ggplot(mapping = aes(x = mean_monthly_tmax_f,
                       y = total)) +
  # points to represent spp total counts and mean monthly max temp
  geom_point(shape = 21,
             color = "darkblue",
             alpha = 0.5) +
  # different panels for each species
  facet_wrap(~ species,
             scales = "free_y",
             ncol = 1) +
  # relabelling axes
  labs(x = "Mean monthly maximum temperature (\U00B0 F)",
       y = "Total counts")

# displaying plot
temp_count_spp
```


# 3. Analyses to examine relationships

If you think one variable _predicts_ another (i.e. you think you could guess at the total counts for a species if you knew a given maximum temperature), you should use a **linear model.**

If you think the variables are related but one variable doesn't necessarily predict another, you should use **correlation**.

There are two types of correlation: **Pearson correlation** and **Spearman correlation**. 










